### 2023-05-23 Manuscript/Draft1.tex
manuscript update

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs. Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.

The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population. A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.

\textbf{use more info from initial proposal}

@@

{\centering \includegraphics[width=0.95\linewidth]{Images/Strata2} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata2}

@@

\caption{Strata inhabited by each species in our population models.}\label{fig:strata-plot}
\end{figure}

@@

{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios}

@@

We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.

@@

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process \textbf{cite?} and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. \textbf{BL: Help with correct notation for adding noise}. The impication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Tabl XXX {[}{[}needs to be added{]}{]} lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.

@@

\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following
7 columns. There are 3*3*2*2*2*2*2 = 288 VAST model combinations and 3*3*2*2*2*2 = 144 stratified mean estimates.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllll}
\toprule
Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Season & Covariates (VAST)\\
\midrule
Yellowtail & Increasing & Repeating & All strata & No & Spring & No\\
Cod & Constant & Increasing 5°C & Subset & Yes & Fall & Yes\\
Haddock & Decreasing &  &  &  &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under the variety of conditions shown in Table \ref{tab:scenarios}. and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata. The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain. More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \ref{fig:PopPct}. This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between abundance estimates and model output for each of our abundance estimates. While the model-based results tended to provide a slightly lower mean absolute error compared to the stratified mean with a mean relative error of 0.34 compared to 0.38, the variance seen in VAST relative errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. As a result, while \(77/80\approx 96\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the high variance in error values indicate how sensitive model-based estimates were to the settings used. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

@@

Including covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. As a result, of the VAST estimates that included covariates information, \(47/48\approx98\)\% had a lower relative error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.

@@

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-azarovitz1981brief}{}}%
Azarovitz, T. 1981. A brief historical review of the woods hole laboratory trawl survey time series. Bottom trawl surveys.

@@

@@

@@

\caption{Percent of each species in each strata for during survey weeks in our spatial simulations. All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right.}\label{fig:PopPct}
\end{figure}

>>>>

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.). Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.

The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.

@@

{\centering \includegraphics[width=0.95\linewidth]{Images/GB_strata} \includegraphics[width=0.95\linewidth]{Images/Strata2} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata2}

@@

\caption{Strata inhabited by each species in our population models. All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange. Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row. Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
\end{figure}

@@

{\centering \includegraphics[width=0.99\linewidth]{Images/Population_scenarios2}

@@

We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.

@@

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref(fig:PopPct). The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species.

@@

\caption{\label{tab:scenarios}Each index estimate chooses one condition from each of the following 7 columns.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllll}
\toprule
Species & Population Trend & Temperature Scenario & Strata Included & Noise Added & Season & Covariates (VAST)\\
\midrule
Yellowtail & Increasing & Repeating & All strata & No & Spring & No\\
Cod & Constant & Increasing 5°C & Subset & Yes & Fall & Yes\\
Haddock & Decreasing &  &  &  &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 28). The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain. More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \ref{fig:PopPct}. This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between abundance estimates and model output for each of our abundance estimates. While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used. When considering individual scenarios, \(77/80\approx 96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations. Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that \(\approx 63\%\) of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate. VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower errors than when covariates are included). We analyze these results further in the sections that follow.

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

@@

Including covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. When comparing to the design-based estimates, \(47/48\approx98\)\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.

@@

As seen in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-arreguin1996catchability}{}}%
Arreguı́n-Sánchez, F. 1996. Catchability: A key parameter for fish stock assessment. Reviews in fish biology and fisheries \textbf{6}: 221--242. Springer.

\leavevmode\vadjust pre{\hypertarget{ref-azarovitz1981brief}{}}%
Azarovitz, T. 1981. A brief historical review of the woods hole laboratory trawl survey time series. Bottom trawl surveys.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Econ}{}}%
{Fisheries Economics of the United States: Data and Visualizations}. (n.d.). \url{https://www.fisheries.noaa.gov/data-tools/fisheries-economics-united-states-data-and-visualizations\#}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-langan2021climate}{}}%
Langan, J.A., Puggioni, G., Oviatt, C.A., Henderson, M.E., and Collie, J.S. 2021. Climate alters the migration phenology of coastal marine species. Marine Ecology Progress Series \textbf{660}: 1--18.

@@

\caption{Percent of each species in each strata for during survey weeks in our spatial simulations. All constant temperature scenarios follow the patterns on the left while increasing temperature scenarios follow the patterns on the right. See Figure 1?? for a spatial reference of the Georges Bank strata.}\label{fig:PopPct}
\end{figure}
